Supplementary Information for RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data

### 2014-03-17 doc/bioinformatics2014/document.tex
Update manuscript. Addressed all the comments from ISMB.

\documentclass{bioinfo}
\copyrightyear{2014}
\pubyear{2014}
\usepackage{amsmath}
\usepackage{afterpage}
\usepackage{algorithm}
\usepackage{bbm}

%\usepackage{placeins}
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{xr}
\externaldocument{supplementary_info}
\algnewcommand{\LineComment}[1]{\State \(\triangleright\) #1}

@@

\section{Motivation:}
Next-generation sequencing technology is increasingly being used for clinical diagnostic tests. Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations. However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.

\section{Results:}
We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency. We identify six mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

@@

\section{Introduction}

Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}. Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}. As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird:2010ab}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}. NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.

@@

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.  VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}. RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}. However, that algorithm requires a very high read depth to estimate the sequencing error rate and call variants.

@@

The remainder of this article is organized as follows. In the next section we describe the statistical model structure of our new algorithm, RVD2. Then, we derive a sampling algorithm for computing the posterior distribution over latent variables in the model and use those samples in a Bayesian posterior distribution hypothesis test to call variants. We compare the performance of RVD2 to several other variant calling algorithms for a range of read depths and minor allele fractions. Finally, we show that RVD2 is able to call variants on a heterogeneous clinical sample and identify two novel loss-of-heterozygosity events.

%%%%%%%%%%%%%%%%%%
% Model Structure
%%%%%%%%%%%%%%%%%%
\section{Model Structure}\label{sec:model_structure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
\caption{RVD2 Graphical Model}
\label{fig:graphical_model}
\end{center}
\end{figure}

@@

Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true. We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.

%%%%%%%%%%%%%%%%%%
% Data Sets
%%%%%%%%%%%%%%%%%%
\section{Data Sets}

@@

We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample. To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j^{\text{control}} \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.2]$, heterozygous mutant $[0.1, 0.9]$, and homozygous mutant $[0.8, 1.0]$ and the size of the test is $\alpha=0.05$.

@@

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance.}
\label{fig:ROC}
\end{center}
\end{figure}

@@

We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity. In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants. The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives. GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.

\subsubsection*{Sensitivity/Specificity Comparison}

\begin{figure*}[!tpb]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_ss.pdf}
\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
\label{tbl:comparison_ss}
\end{center}
\end{figure*}

@@

Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance. They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success. GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples. Therefore, those algorithms are expected to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples. VarScan2-somatic is able to call more mixed samples. However, as the read depth increases the specificity degrades. Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below. muTect has good performance across a wide range of MAF levels. But even at the highest depth only has around 0.5 sensitivity for low MAF levels.

The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample. The sensitivity increases considerably with read depth at a slight expense to specificity. With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs. However, in practice one may not know the optimal $\tau^*$ a-priori. With $N=6$ replicates, the sensitivity increases considerably for low MAF variants with a slight degradation in specificity due to false positives. When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.

\subsubsection*{False Discovery Rate Comparison}
Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, strelka and muTect. Blank cells indicate no positive calls were made.

@@

\begin{figure*}[!tpb]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_fdr.pdf}
\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
\label{tbl:comparison_fdr}
\end{center}
\end{figure*}
% from folder 'rvd2\results\2013-09-19_operating_characteristics'
%\afterpage{\clearpage}

VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level. However, the FDR is high due to many false positives. Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth. Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF. However, it does not call any variants at or below 1\% MAF.  MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths. But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\% -- 100\% MAF.

@@

\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}

\begin{figure*}[!tpb]
\begin{center}
\includegraphics[width=0.9\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}}$  is significantly different from $ \mu^{\text{control}}$.}
\label{fig:brca_MAF}
\end{center}
\end{figure*}
%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
%TODO: add star to test that have mudelta > .16 Done.
%TODO: set tau=0.16 for mudelta test.  Done

RVD2 identified six variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. Of the six called variants, three were homozygous germline variants and two were heterozygous germline variants. The two heterozygous variants had a loss-of-heterozygosity in the tumor sample. The remaining one is a homozygous somatic variant. Because the median read depth for the case sample is 60, we set the threshold of the test for $\mu^\Delta$ to $\tau = 16\%$ to minimize the false discovery rate (Figure~\ref{tbl:comparison_fdr}).

Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C are called significantly different in tumor and normal sample MAF. Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations. Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample. In the tumor sample we identified six mutations chr7:154743899C$>$T, chr7:154753635T$>$C, \\ chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C and chr7:154780960C$>$T. Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events. Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding identities are shown in the figure.
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

The original research describing this sample used Strelka to identify mutations in the same sample. They identified chr7:154760439 as variant, but did not call the other five variants. In particular strelka missed the two LOH events.  % To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample. Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.

@@

Our inference algorithm uses Gibbs sampling to estimate the RVD2 hierarchical empirical Bayes model. This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically. However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches. We opted for this balance of speed and accuracy because computational time is often not limiting and the cost of a false positive or false negative greatly outweighs the cost of more computation.

>>>>

\documentclass{bioinfo}
\copyrightyear{2014}
\pubyear{2014}
\usepackage{amsmath}
\usepackage{afterpage}
\usepackage{algorithm}
\usepackage{bbm}
\usepackage{subfig}
%\usepackage[font={small}]{caption}
\usepackage[skip=2pt,font=scriptsize]{caption}
%\usepackage{placeins}
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{xr}
\externaldocument{supplementary_info}
\algnewcommand{\LineComment}[1]{\State \(\triangleright\) #1}

@@

\section{Motivation:}Next-generation sequencing technology is increasingly being used for clinical diagnostic tests. Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations. However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.

%TODO Update the chinical dataset result
\section{Results:}We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency. We identify six mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

@@

\section{Introduction}
\label{sec:intro}
Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}. Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}. As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird:2010ab}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}. NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.

@@

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples. Some approaches attempt to estimate the tumor sample purity then adjust the calling algorithm with that estimation.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP. However, RVD2 and some other algorithms are able to call variations in heterogeneous samples without estimating the tumor purity. VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}. RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}. However, that algorithm requires a very high read depth to estimate the sequencing error rate and call variants.

@@

The remainder of this article is organized as follows. In the next section we describe the statistical model structure of our new algorithm, RVD2. Then, we derive a sampling algorithm for computing the posterior distribution over latent variables in the model and use those samples in a Bayesian posterior distribution hypothesis test to call variants. We compare the performance of RVD2 to several other variant calling algorithms for a range of read depths and minor allele fractions. Finally, we show that RVD2 is able to call variants on a heterogeneous clinical sample and identify two novel loss-of-heterozygosity events. 
%TODO: Fixed this part
%%%%%%%%%%%%%%%%%%
% Model Structure
%%%%%%%%%%%%%%%%%%
\section{Model Structure}\label{sec:model_structure}

@@

\begin{figure}[!bpth]
\begin{center}
\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
\caption{RVD2 Graphical Model.}
\label{fig:graphical_model}
\end{center}
\end{figure}

@@

Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true. We use the Benjamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
%TODO: Bonferroni correction
%%%%%%%%%%%%%%%%%%
% Data Sets
%%%%%%%%%%%%%%%%%%
\section{Data Sets}

@@

%TODO: change the result
We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample. To identify germline mutations, we computed the empirical posterior mean $ \bar{\mu}_j^{control} $ in the control data and test which interval it falls in, where the intervals are: homozygous reference $[0, 0.05)$, heterozygous mutant $[0.05, 0.95)$, and homozygous mutant $[0.95, 1.0]$. We set the threshold between homozygous reference and heterozygous mutant at a relatively high 0.05 considering the low average read depth (40x for the normal sample and 90x for the tumor sample), likewise for the border between heterozygous and homozygous mutant at 0.95. We reported the 95\% credible interval to shown the dominant region for $ \hat{\mu}_j^{control} $ in called positions afterwards.

@@

$ \chi^2 $ test was performed with p-value set at 0.05 after posterior distribution test. In the synthetic data we used Bejamini-Hochberg method to control the family-wise error rate over positions called by posterior distribution test. However, we didn't apply Benjamini-Hochberg method to clinical HCC1187 sequence data as Benjamini-Hochberg method is well-recognized very conservative, and it restraints RVD2 from detecting low rate mutated positions when read depth is as low as 90x.

%TODO: No bonferoni correction for the chi2 test

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance. Each subfigure shows ROC curves across four different read depths for one MAF level. Within one subfigure it can be seen that the performance improves monotonically with read depth. Across different subfigures, it can be seen that the performance improves with MAF level.}
\label{fig:ROC}
\end{center}
\end{figure}

@@

We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity. Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2 Somatic, Strelka and muTect are designed to call variants in heterogeneous samples, which serve as better comparison to RVD2. In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants. The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives. GATK, Varscan2, Strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.

@@

% from folder 'rvd2\results\2013-09-19_operating_characteristics'
%\afterpage{\clearpage}
\subsubsection*{Sensitivity/Specificity Comparison}
\begin{figure*}[!htpb]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_ss.pdf}
\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms using synthetic sequence data.}
\label{tbl:comparison_ss}
\end{center}
\end{figure*}

\begin{figure*}[!tbhp]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_fdr.pdf}
\caption{False discovery rate comparison of RVD2 with other variant calling algorithms using synthetic sequence data. Blank cells indicate no locations were called variant.}
\label{tbl:comparison_fdr}
\end{center}
\end{figure*}
Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance. They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success. GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples. Therefore, those algorithms are expected to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples. VarScan2-somatic is able to call more mixed samples. However, as the read depth increases the specificity degrades. Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below. muTect has good performance across a wide range of MAF levels. But even at the highest depth only has around 0.5 sensitivity for low MAF levels.

The statistics for RVD is an average from six replicates as RVD uses six control replicates to estimate the model but returns six sets of statistics separately for six case replicates.

RVD uses multiple control replicates to estimate the model but only one pair-end case sample (two replicates) to do hypothesis testing. Given a data set of three pair-end replicates for control sample and three pair-end replicates for case sample, three set of mutation calling results are returned. The statistics for RVD shown in Figure~\ref{tbl:comparison_ss} and Figure~\ref{tbl:comparison_fdr} is an average from three set of statistics. RVD performed the best among all algorithms when the read depth is as high as 40000x. RVD correctly called all the mutated positions across all MAF levels with no false positives when MAF level is 0.3\% or lower. RVD tends to be a little bit over sensitive when the MAF level is 1.0\% or higher, which causes specificity slightly lower than 1.00. However, RVD program fails when the depth is at 5000x or lower and is not able to call any mutations.

The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample. The sensitivity increases considerably with read depth at a slight expense to specificity. With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs. However, in practice one may not know the optimal $\tau^*$ a-priori. With $N=6$ replicates, the sensitivity increases considerably for low MAF variants with a slight degradation in specificity due to false positives. When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
\begin{figure*}[!bpth]
    \centering
    \subfloat[\label{fig:brca_MAF:a}]{{\includegraphics[width=0.65\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf} }}%
    %
    \subfloat[ \label{fig:brca_MAF:b}]{{\includegraphics[width=0.3\textwidth]{pdf_figs/RVD2calltable.pdf} }}%
    \caption{Germline and Somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. HG19 Genomic locations are chr7:154,700,000+X. (a) Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$. The vertical lines represent 95\% credible interval around mean MAF. (b) Mutation type classified by RVD2. RVD2 identified twelve variants in the 44kbp PAXIP1 gene. There were eight germline variants and eight somatic variants. Especially, three heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample. The dbSNPv138 column shows the identities for positions found to be common populations SNPs according to dbSNPv138.}%
    \label{fig:brca_MAF}%
\end{figure*}
\subsubsection*{False Discovery Rate Comparison}

Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, Strelka and muTect. Blank cells indicate no positive calls were made.

@@

VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level. However, the FDR is high due to many false positives. Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth. Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF. However, it does not call any variants at or below 1\% MAF.  muTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths. But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\% -- 100\% MAF. RVD has best FDR performance in the high read depth for 0.1\% -- 1\% MAF levels. The FDR increases to around 0.3 for 10\% -- 100\% MAF in the high read depth.

@@

\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
\subsubsection{Performance of RVD2}

\begin{figure*}[!pthb]
    \centering
    \subfloat[\label{fig:brca_dot:a}]{{\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
    %
    \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
    \caption{(a)Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. VarScan2-somatic reported 60 positions, muTect 11 positions, RVD2 12 positions and Strelka only 1 position. This figure uses position index to show the correspondence of positions called by different algorithms for notation simplicity. Complete actual positions and depth distribution are provided in Supplementary Table 1 for validation. (b) Depth distribution for positions called by RVD2 and muTect.}%
    \label{fig:brca_dot}%
\end{figure*}

RVD2 identified twelve variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. There were eight germline variants and eight somatic mutations in the twelve variants. Especially, three heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.

Figure~\ref{fig:brca_MAF:a} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154743899 C$>$T, chr7:154753635 T$>$C and chr7:154780960 C$>$T were called germline homozygous mutations. Positions chr7:154749704, chr7:154754371 and chr7:154758813, chr7:154766700, chr7:154781769 were called heterozygous in the normal sample. Positions chr7:154749704, chr7:154753635 T$>$C, chr7:154754371 T$>$C, chr7:154758813 G$>$A, chr7:154760439 A$>$C, chr7:154766732,  chr7:154766832, chr7:154777118 were called significantly different in tumor and normal sample MAF. Positions chr7:154749704, chr7:154754371 T$>$C and chr7:154758813 G$>$A were called loss-of-heterozygosity events. Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding identities are shown in the Figure~\ref{fig:brca_MAF:b}. The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1 for specificity validation. 
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

\subsubsection{Performance comparison with other algorithms}

For further detection validation and performance comparison, we ran muTect and VarScan2-somatic to call mutations in the PAXIP1 gene in HCC1187 sample. We also compared to the result shown in original research report where Strelka was used to identify mutations in the same sample ~\citep{hcc1187techreport}. Figure~\ref{fig:brca_dot:a} shows mutation detection result from Strelka, RVD2, muTect, and VarScan2-somatic, the state-of-art algorithms able to call mutation from heterogeneous samples. For notation simplicity, we use position index to present actual positions in Figure~\ref{fig:brca_dot}, while the correspondence is provided in Supplementary Table 1.

The mutations called by RVD2 and muTect are the most consistent among all the techniques. RVD2 detected twelve germline and somatic mutations, while muTect reported eleven, ten in common. In the disagreements, RVD2 did not call position 50 while muTect did not call position 37 and 58. Referring to the depth distribution shown in Figure~\ref{fig:brca_dot:b}, it can be seen that position 37 and 58 are more likely mutated while position 50 is less likely mutated.

Strelka was the least sensitive algorithms among all the algorithms. According to the technical report, Strelka identified position 22 (chr7:154760439) as variant, but did not call any other variants. In particular Strelka missed the three LOH events called by RVD2. On the contrary, VarScan2-somatic was the most sensitive algorithms, and called sixty positions as shown in Figure~\ref{fig:brca_dot:a}. VarScan2-somatic  detected all the positions called by RVD2 except position 35, which turns out to be a very likely mutation given the depth distribution in Figure~\ref{fig:brca_dot:b}. This fact reveals the imperfect though high sensitivity for VarScan2-somatic. On the other side, VarScan2-somatic reported fifty positions which were not called by any other three algorithms. The read depth in Supplementary Table 1 suggestions that these positions are very likely to be false positives. The fact that VarScan2-somatic can be over-sensitive has appeared in synthetic dataset analysis. As shown in Figure~\ref{tbl:comparison_fdr}, the False Discovery Rate for VarScan2-somatic at read depth 53 MAF level 1.0\% is as high as 1.00.  \citet{spencer2013performance} also mentioned that VarScan2 has tendency to be over-sensitive at high read depth. 
% To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample. Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.

@@

Our inference algorithm uses Gibbs sampling to estimate the RVD2 hierarchical empirical Bayes model. This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically. However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches. We opted for this balance of speed and accuracy because computational time is often not limiting and the cost of a false positive or false negative greatly outweighs the cost of more computation. Another factor that can affect the speed of RVD2 is the number of Metropolis-Hasting sample within one Gibbs sampling run. However, RVD2 is able to use N cores in parallel ( no more than the number of Metropolis-Hasting samples), which can significantly decrease the running time to $ \dfrac{1}{N} $ of the running time in sequence. Finally, the running time for RVD2 is nearly linearly proportional to sequence length of the sample. Taking our 400bp long synthetic data as an example, it takes around 1.5 seconds when we run the program in sequence with 10 Metropolis-Hasting samples in one Gibbs sample. Therefore, if we take 4000 Gibbs samples, it takes around $ \dfrac{1.5\times4000}{60} = 100 $mins to run the program. If we use 10 cores for parallel processing, it will take around $ \dfrac{100}{10}=10 $mins to run the program. Considering the running time for RVD2 can be much longer for long sequences, we are currently developing a variational method to estimate RVD2 model.

### 2014-03-17 doc/bioinformatics2014/supplementary_info.tex
Update manuscript. Addressed all the comments from ISMB.

\documentclass[11pt,reqno]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{algorithm, algorithmic}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pdfpages}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

\title[RVD2: Supplementary Information]{Supplementary Information for
  RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
%\date{}                                           % Activate to display a given date or no date

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
\label{fig:depthM}
\end{center}
\end{figure}

@@

\section{Histogram of $ \mu_j $ for mutations detected by RVD2 in HCC1187 dataset}
\begin{figure}[H]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
\label{fig:hist_pos}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_neg.pdf}
\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
\label{fig:hist_neg}
\end{center}
\end{figure}

\section{Comparison between RVD2, VarScan2 somatic and muTect in Clincal HCC1187 dataset}
\includepdf[pages={1-},scale=0.9]{pdf_figs/position_lookup_chart_all.pdf}

>>>>

\documentclass[11pt,reqno]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{algorithm, algorithmic}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pdfpages}
\usepackage[skip=2pt,font=scriptsize]{caption}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

\title[RVD2: Supplementary Information]{Supplementary Information for
  RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data}
%\date{}      % Activate to display a given date or no date

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.85\textwidth]{pdf_figs/depthM.pdf}
\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
\label{fig:depthM}
\end{center}
\end{figure}

@@

\section{histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 44kbp paxip1 gene from chr7:154738059 to chr7:154782774 in hcc1187 dataset}
\begin{figure}[H]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly higher than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j> \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.} 
\label{fig:hist_pos}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_neg.pdf}
\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly lower than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j< \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
\label{fig:hist_neg}
\end{center}
\end{figure}

\bigskip
\bigskip

\includepdf[pages={1-},scale=1]{pdf_figs/position_lookup_chart_all.pdf}
